Unsupervised learning of Gaussian mixture models in linear time using a variational Bayesian method

### 2017-05-17 articles/nips-2017/ms.tex
spell out the problematic terms

Consider that the number of true components $K_{true}$ is unknown, and that we
have an optimal mixture of $K$ components in the sense that the model
parameters $\vectheta$ minimise the total message length in Eq.
\ref{eq:objective-function}.  We are interested to know whether the total
message length will decrease if we add or remove components.  For
example, our current state may be an optimized mixture with $K=2$ and we want
to know if an optimal mixture with $K=3$ will result in a shorter message
length.  By denoting $I\current$ as the message length of the current (optimal) 
mixture and $I\future$ as mixtures that have yet to be computed, we
want to know if $I\future - I\current < 0$, or rather we seek to calculate 
some probability $\textrm{Pr}(I\future - I\current < 0)$ before having to 
optimize the future mixture $I\future$.  
Taking Eq. \ref{eq:objective-function}, we can derive an expression for the 
difference in message lengths between future mixtures and the current state:
\begin{eqnarray}
  I\future - I\current &=& \Delta{K}\left[
  \left(1 - \frac{D}{2}\right)\log{2} + \frac{1}{4}\left(D(D+3)+2\right)\log\frac{N}{2\pi}\right] \nonumber \\
  &-& \log\Gamma(K+\Delta{K}) + \log\Gamma(K) + \frac{1}{2}\left(\log{Q\future} - \log{Q\current}\right) \nonumber \\
  &+& \frac{1}{2}\left(\frac{D(D+3)}{2} - 1\right)\left(\sum_{k=1}^{K+\Delta{K}}\log\weight_k\future - \sum_{k=1}^{K}\weight_k\current\right) \nonumber \\
  &-& \sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future + \sum_{n=1}^{N}\log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\current \nonumber \\
  &+& \frac{D + 2}{2}\left(\sum_{k=1}^{K}\log|\veccov_k|\current - \sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future\right) \quad .
\end{eqnarray}

>>>>

Consider that the number of true components $K_{true}$ is unknown, and that we
have an optimal mixture of $K$ components in the sense that the model
parameters $\vectheta$ minimise the total message length in Eq.
\ref{eq:objective-function}.  We are interested to know whether the total
message length will decrease if we add or remove components.  For
example, our current state may be an optimized mixture with $K=2$ and we want
to know if an optimal mixture with $K=3$ will result in a shorter message
length.  By denoting $I\current$ as the message length of the current (optimal) 
mixture and $I\future$ as mixtures that have yet to be computed, we
want to know if $I\future - I\current < 0$, or rather we seek to calculate 
some probability $\textrm{Pr}(I\future - I\current < 0)$ before having to 
optimize the future mixture $I\future$.  
Taking Eq. \ref{eq:objective-function}, we can derive an expression for the 
difference in message lengths between future mixtures and the current state:
\begin{eqnarray}
  I\future - I\current &=& \Delta{K}\left[
  \left(1 - \frac{D}{2}\right)\log{2} + \frac{1}{4}\left(D(D+3)+2\right)\log\frac{N}{2\pi}\right] \nonumber \\
  &-& \log\Gamma(K+\Delta{K}) + \log\Gamma(K) + \frac{1}{2}\left(\log{Q\future} - \log{Q\current}\right) \nonumber \\
  &+& \frac{1}{2}\left(\frac{D(D+3)}{2} - 1\right)\left(\sum_{k=1}^{K+\Delta{K}}\log\weight_k\future - \sum_{k=1}^{K}\weight_k\current\right) \nonumber \\
  &-& \sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future + \sum_{n=1}^{N}\log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\current \nonumber \\
  &+& \frac{D + 2}{2}\left(\sum_{k=1}^{K}\log|\veccov_k|\current - \sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future\right) \quad .
  \label{eq:delta_I}
\end{eqnarray}

Although there are many terms in Eq. \ref{eq:delta_I}, most are readily
available and can be immediately computed for given trials of $\Delta{K}$.  
Only three terms are exceptions, which expectedly relate to the parameter 
estimates of future, uncomputed mixtures:
\begin{enumerate}
  \item The sum of the log of the weights of future mixtures,\\
        $\sum_{k=1}^{K+\Delta{K}}\log\weight_k\future$.
  \item The negative log-likelihood of future mixtures,\\ $-\sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future$.
  \item The negative sum of the log of the determinant of covariance matrices
        of future mixtures,\\
        $-\sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future$.
\end{enumerate}
